Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2020-12-22 jss/EpiCompare-paper/EpiCompare-paper.tex
added an updated section about r0 and time invariant analysis

@@

The current epidemic of COVID-19 also highlights that infectious disease
models are only one piece of the overall analysis pipeline. University
based resources like John Hopkin's and government numerical dashboards
(across all levels of government) during the COVID-19 epidemic remind us
that descriptive statistics and visualization can be a important first
step in the process (multiple \cite{}?).\}
\textcolor{orange}{add comment about NYT or something} Still, rightly
so, a large amount of theoretical work goes into modeling epidemics,
with different models focusing at the individual / agent level, network
structure or just aggregate flows (review paper \cite{}?). All placing
individuals / proportions of the populations into different states
(e.g.~suspectible, exposed, infected, recovered, etc.). With all these
models, review and comparison papers in the literature and through MIDAS
(Models of Infectious Disease Agent Study) Control Center helps the
individual practitioner decide the correct approach.
\textcolor{orange}{along with expertise from healthcare professionals...}

@@

\subsection[r0]{Motivation through \(R_0\)}\label{r0}

\(R_0\), the initial reproduction number, has been called the ``most
important quantity in epdiemiology'' \citep[][]{Gallagher2020}. To see
the importance of \(R_0\), one need only read the newspaper
\citep{Fisher2020} or look at the length of the table of estimated
\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
maybe, the most famous \textit{time-invariant} numerical summary of an
epidemic, and is commonly associated with the
Susceptible-Infectious-Recovered (SIR) data/models. The SIR framework,
composed of data and statistical models where individuals pass from
Susceptible to Infectious to Recovered states is a common, if not the
most common, modeling framework in infectious disease epidemiology (seen
in examples of recent published works in \cite{FILLIN}).

In regards to traditional \(X\) vs.~time plots, \(R_0\) is difficult to
visualize. For example, in a SIR simulation with two models have the
same value of \(R_0\), one can see in Fig.
\ref{fig:different-scales-standard} that there is no way of knowing that
from looking at the graph (this figure will be commented on more below).

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-standard.pdf}%\includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Bivariate view of \# in each state vs. time. Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$.\textcolor{violet}{We probably need to define $\beta/\gamma$.}
\textcolor{violet}{NEW: recommendation - make this plot actually have 3 examples (the current one), one where we have an affine transformation of the time scale, and one with the halleolof data.} \textcolor{violet}{Also these figures should be reproducible.}}
    \label{fig:different-scales-standard}
\end{figure}

\subsection[ternary visualizations]{Ternary and time-invariant
visualizations for \(R_0\)}\label{sec:ternary}

It is possible to identify, at a glance, which SIR epidemic paths, like
those in Fig. \ref{fig:different-scales-tern} have the same value of
\(R_0\) via ``time-invariant'' visualizations, under some circumstances.
Specifically, for SIR models we propose using ternary to examine an
epidemic's trajectory in a more time-invariant manner.

Ternary plots are sometimes used in chemistry \citep[][]{Gillespie1976}
but only rarely seen in the field of infectious disease epidemiology (to
our knowledge, only seen in \cite{}). Ternary plots illustrate the
relationship of constrained 3D data, namely of points
\((a_i, b_i, c_i) \in [0,1] \times [0,1] \times [0,1]\) with
\(a_i+ b_i + c_i =1\) and as such, are situational. The SIR framework
happens to be of the situational format required for ternary plots where
\((S(t), I(t), R(t))\) are the number of susceptible, infectious, and
recovered individuals in each state at time \(t\) and
\(S(t) + I(t) + R(t) = N(t)\) where \(N(t)\) is the population size at
time \(t\).

Ternary plots are time-invariant in the sense that the temporal scale
does not explicitly apear in the visualization and as a consequence,
allow for direct comparisons of outbreaks on different time scales
(e.g.~days vs.~years) or of SIR data of the same disease in different
areas (our first case study in Sec. \ref{sec:ex1} focuses on this).

To show that ternary plots have to potential to aid in the comparison of
different epidemic's \(R_0\)s -in certain circumstances- we first need
to introduce the classic, deterministic Kermack and McKendrick
\cite{CITETHIS?} SIR model where the initial number in each state
\((S(0), I(0), R(0))\) are known and \(N\) the population size is
constant. The movement of individuals among the states is given by the
following differential equations (Eq. \eqref{eq:sir-ode}) where
\(\beta\) is the average infection rate and \(\gamma\) is the average
recovery rate, \begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align}

\textcolor{violet}{SUGGESTION: define $R_0$ mathematically here (though referred to in figure 2 - which still needs to be defined there).}

If we have two SIR models that follow the Kermack and McKendrick SIR
model, have the same percentage of individuals in the initial states,
and the same value of \(R_0\) then the two models will create exactly
overlapping paths in a ternary plot.

@@

\begin{figure}[!ht]
%% Code to reproduce is in inst/paper_figs/different-time-scales.R
    \centering
    \includegraphics[width = .5\textwidth]{images/diff-time-ternary.pdf}
    \caption{Ternary view of \# in each state.  Both this Fig. and Fig. \ref{fig:different-scales-standard} display the same two sets of data.  Both data sets are generated using the same value of $R_0 = 2.8$ but have different values of $\beta$ and $\gamma$. \textcolor{violet}{We probably need to define $\beta/\gamma$.}  While there are obvious are differences in Fig.\ref{fig:different-scale-standard}, the data sets look quite similar in the ternary view. \textcolor{violet}{I might suggest using opacity to show time scale here?} \textcolor{violet}{Make this reproducible too?}}
    \label{fig:different-scales-tern}
\end{figure}

In \pkg{EpiCompare}, we also provide visualization tools to aid in
comparing models in the ternary / time-invariant space, which is
presented in more details in Section
\ref{sec:time-invariant-in-simplex}.

\subsection[Simplexes beyond ternary plots]{Simplexes beyond ternary
plots and time-invariant comparison
tools}\label{sec:time-invariant-in-simplex}

@@

{\centering \includegraphics{Figs/unnamed-chunk-6-1}

@@

\caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.  We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-6}
\end{figure}
\end{CodeChunk}

@@

@@

\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.  We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-8}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-9-1}

@@

\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.  The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-9}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-16-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-16}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0036733 & 0\\
EpiModel SIR & 0.0028813 & 0\\
\bottomrule
\end{tabular}
\end{table}

>>>>

% Pandoc citation processing

@@

The current epidemic of COVID-19 also highlights that infectious disease
models are only one piece of the overall analysis pipeline. University
based resources like John Hopkins' and government numerical dashboards
(across all levels of government) during the COVID-19 epidemic remind us
that descriptive statistics and visualization can be a important first
step in the process (multiple \cite{}?).\}
\textcolor{orange}{add comment about NYT or something} Still, rightly
so, a large amount of theoretical work goes into modeling epidemics,
with different models focusing at the individual / agent level, network
structure or just aggregate flows (review paper \cite{}?). All placing
individuals / proportions of the populations into different states
(e.g.~suspectible, exposed, infected, recovered, etc.). With all these
models, review and comparison papers in the literature and through MIDAS
(Models of Infectious Disease Agent Study) Control Center helps the
individual practitioner decide the correct approach.
\textcolor{orange}{along with expertise from healthcare professionals...}

@@

\subsection[r0]{What \(R_0\) has to do with time-invariant
analysis}\label{r0}

\(R_0\), the initial reproduction number, has been called the ``most
important quantity in epdiemiology'\,' \citep[][]{heesterbeek2003}. By
definition, \(R_0\) is the number of expected secondary infections when
a primary infection is introduced to a susceptible population. To see
the importance of \(R_0\), one need only read the newspaper
\citep{Fisher2020} or look at the length of the table of estimated
\(R_0\) quantities for COVID-19 \citep{Aronson2020}. \(R_0\) is also,
maybe, the most famous \textit{time-invariant} numerical summary of an
epidemic, which allows epidemics to be compared to one another in both
different time and geographic scales.

Estimator(s) for \(R_0\) are dependent on the epidemic modeling
framework, which consists of which states an individual can occupy
(e.g.~susceptible, infectious, recovered) and a description of how
individuals move from one state to the next over time. A common epidemic
modeling framework is the SIR model, originally introduced by
\citet{kermack1927}. Transitions from one state to the next are defined
by a series of ordinary differential equations, where \(N\) is the
(fixed) total number individuals, \(\beta\) is the rate of infection,
and \(\gamma\) is rate of recovery, \begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align} From this,
\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
estimated infection rate compared to the estimated recovery rate.

In regards to traditional \(state\) vs.~time plots, \(R_0\) is difficult
to visualize, especially with respect from one epidemic to another. For
example, consider the scenarios where the first epidemic is generated
from a SIR model with \((S(0) = 990, I(0) = 10)\), \(\beta_1 = .3\) and
\(\gamma_1 = .15\), and the second epidemic is generated from a SIR
model with \((S(0) = 990, I(0) = 10)\), \(\beta_2 = .24\) and
\(\gamma_1 = .12\) over 40 days. Both epidemics have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:sir-ex}. At a glance, we may assume that Model 1 has a larger
\(R_0\) than Model 2 because the peak of infection occurs more quickly
than in Model 2. On the other hand, we may think Model 2 has a larger
\(R_0\) because it has a slightly larger peak of infection than Model 1.

\begin{CodeChunk}
\begin{figure}[H]

{\centering \includegraphics{Figs/unnamed-chunk-2-1}

\caption{\label{fig:sir-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.}\label{fig:unnamed-chunk-2}
\end{figure}
\end{CodeChunk}

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig. \ref{fig:ternary-ex}, the
points seem to overlap and form the same trajectory. In other words, we
can see there is something fundamentally linking these two different
epidemics, and this fundamental link turns out to be \(R_0\).

@@

\begin{CodeChunk}
\begin{figure}[H]

{\centering \includegraphics{Figs/unnamed-chunk-3-1}

\caption{\label{fig:ternary-ex}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\subsection[Simplexes beyond ternary plots]{Simplices beyond ternary
plots and time-invariant comparison
tools}\label{sec:time-invariant-in-simplex}

@@

@@

\caption{\label{fig:cifs}Empirical cumulative incidence functions of prodrome (symptom) onset and measles rash appearance.  We see that there is approximately a a constant lag between the two curves.}\label{fig:unnamed-chunk-8}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-10-1}

@@

\caption{\label{fig:hag-tern-raw}Time invariant view of the Hagelloch epidemic where we view the individuals in Susceptible, Infectious, or Recovered states.  We see there are two peaks of infection (the vertical axis).}\label{fig:unnamed-chunk-10}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-11-1}

@@

\caption{\label{fig:tern-class-data}Time invariant outbreak curves for the three class groups.  The pre-school class has a distinct peak of infection whereas the peak infection point for the other two classes are less well defined.}\label{fig:unnamed-chunk-11}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-18-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-18}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0036733 & 0.00\\
EpiModel SIR & 0.0149686 & 0.02\\
\bottomrule
\end{tabular}
\end{table}
